Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-02 talk/icooolps2009/paper.tex
another round of reading

\fijal{I'm lost, will continue later}
A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is its bytecode dispatch loop. Usually that is is also the only hot
loop of the language interpreter \arigo{Uh?}.  Tracing one iteration of this loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.

@@

Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes and all register and the
accumulator are integers. A simple program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

The condition for reusing already existing machine code needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. They correspond to different
unrollings and paths of that loop though. To figure out which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be entered. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

Let's look at which hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. The basic thing needed to apply hints is a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red.

>>>>

\fijal{I find following paragraph out of scope and completely confusing, we
should instead simply state that we unroll the loop, how we do that and
why we do that. Completely ignore aspect of an interpreter loop I suppose,
because everything previously keeps talking about can\_enter\_jit that closes
loop being available at jump back bytecodes}
A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is its bytecode dispatch loop. Usually that is is also the only hot
loop of the language interpreter \arigo{Uh?}.  Tracing one iteration of this loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.

@@

\fijal{This paragraph should go away as well}
Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes and all register and the
accumulator are integers. A simple program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

\fijal{This is wrong. Without virtuals there is also at most one assembler
loop per user loop. If it has more branches, we enter the loop as usual and
then we create a bridge for a new situation}
The condition for reusing already existing machine code needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. They correspond to different
unrollings and paths of that loop though. To figure out which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be entered. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

\fijal{Stopped reading at that point}
Let's look at which hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. The basic thing needed to apply hints is a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red.
